Changes in Mobility and Site Occupation during the Late Pleistocene in Korea

### 2020-01-08 analysis/paper/paper.Rmd
Editing based on Ben's comments

The appearance of stemmed points in the Late Pleistocene (~40-35 ka) in Korea is often assumed to have transformed forager lifeways because it is thought to reflect more specialed hunting practices (citations). Stemmed points are likely to have been projectile weapon tips because the stem of these tools is cconsidered to represent halfting technology that joins the lithic to a wooden shaft. Stemmed points represent the first appearance of halfting technology in Korea. Here we explore questions about these transformations using as behavioural ecology framework to explore changes in the way people used the landscape, and the way people used sites. We examined stone artifacts from 22 sites in South Korea to investigate the technological transition represented by the appearance of a new tool, stemmed points, and the blade industry that followed. Previous work on this technological change has focussed on issues of the transition from the Early to the Late Paleolithic and the disperal of _Homo sapiens_ (citations). To our knowledge, the implications for forager mobility and site use have not been explored in detail. Most Korean Paleolithic sites do not preserve organic remains, so we lack direct faunal evidence on hunting and subsistence strategies. Here, we attempt to identify patterns of mobility through analysis of the stone artefact assemblages using Human Behavioral Ecology (HBE) models. Drawing on established relationships between lithic technology and forager mobility (citations, Shott, Kuhn) we hypothesize that the appearance of the new hunting tool might reflect a preference for more portable and efficient technologies, that are part of a broader strategy of moving frequently and further, possibly as an adaptation to population or environmental changes.

<!-- sgenerally, how do we make sense of new stone tools... When a new stone artefact technology appears, how do we explain that? new species... population movements... (great basin)... environment... (Mesolithic & bow and arrow)...  risk management (Australian small tool tradition) -->

@@

In the Eastern hemisphere, striking technological innovations during the Late Pleistocene are restricted to Northeast Asia, such as northern China, Japan and Korea. These innovations include blade and micro technology, high frequencies of retouched blade tools, several novel refined tools such as projectile points, stemmed points, end scrapers, burins, denticulates, etc, and using high-quality raw materials [@bar1999big; @brantingham2001initial; @Bar-Yosef_2002; @bae2017origin; @nakazawa2018quaternary]. Previous work has argued that the appearance of new stone artefact technologies in this region may be linked to modern human dispersals (citation).

The Shuidonggou site in northern China is a typical site for showing the transition from cobble to blade tool industries in the Late Pleistocene of Northeast Asia [@Brantingham_Perreault_2010; @Gao_Zhang_Yang_Shen_Wu_2010; @pei2012shuidonggou]. Shuidonggou Localities SDG 2, 7-12 have blades and bladelets made of higher quality or exotic raw materials in addition to more than 80 polished ostrich egg-shell beads (SDG 2), bone needles and an awl (SDG 12) (citation). The blade compoenent is claimed to have come from Siberia and/or Mongolia at 41 ka, and at 10.5 ka  microblade technology appears but the origin is uncertain. <!-- expand on ages of all the things, comment about the cause of the new things at this site, check, https://link.springer.com/article/10.1007/s10963-019-09129-w  what is the debate? -->

>>>>

The appearance of stemmed points in the Late Pleistocene (~40-35 ka) in Korea is often assumed to have transformed forager lifeways because it is thought to reflect more specialed hunting practices [@lee2002new;@seong2008tanged;@chang2013human]. Stemmed points are likely to have been projectile weapon tips because the stem of these tools is cconsidered to represent halfting technology that joins the lithic to a wooden shaft. Stemmed points represent the first appearance of halfting technology in Korea. Here we explore questions about these transformations using as behavioural ecology framework to explore changes in the way people used the landscape, and the way people used sites. We examined stone artifacts from 22 sites in South Korea to investigate the technological transition represented by the appearance of a new tool, stemmed points, and the blade industry that followed. Previous work on this technological change in Korea has focussed on issues of the transition from the Early to the Late Paleolithic and modern human dispersals [@norton2009evolution; @bae2010origin]. To our knowledge, the implications for forager mobility and site use have not been explored in detail. Most Korean Paleolithic sites do not preserve organic remains, so we lack direct faunal evidence on hunting and subsistence strategies. Here, we attempt to identify patterns of mobility through analysis of the stone artefact assemblages using Human Behavioral Ecology (HBE) models. Drawing on established relationships between lithic technology and forager mobility [@shott1986settlement; @Hiscock_1994; @kuhn1994formal; @kuhn2004upper; @kuhn2015artifacts; @kuhn2016moves; @capriles2018], we hypothesize that the appearance of the new hunting tool might reflect a preference for more portable and efficient technologies, that are part of a broader strategy of moving frequently and further, possibly as an adaptation to population or environmental changes.

<!-- generally, how do we make sense of new stone tools... When a new stone artefact technology appears, how do we explain that? new species... population movements... (great basin)... environment... (Mesolithic & bow and arrow)...  risk management (Australian small tool tradition) --> 
<!-- Do you want to have one more paragraph for the general tradition of explaining the emergence of new stone tools? -->

@@

In the Eastern hemisphere, striking technological innovations during the Late Pleistocene are restricted to Northeast Asia, such as northern China, Japan and Korea. These innovations include blade and micro technology, high frequencies of retouched blade tools, several novel refined tools such as projectile points, stemmed points, end scrapers, burins, denticulates, etc, and using high-quality raw materials [@bar1999big; @brantingham2001initial; @Bar-Yosef_2002; @bae2017origin; @nakazawa2018quaternary]. Previous work has argued that the appearance of new stone artefact technologies in this region may be linked to modern human dispersals by focusing on the origin of the new tools [@seong2008tanged;@bae2010origin;@Bae_2010; @bae2017origin].

The Shuidonggou site in northern China is a typical site for showing the transition from cobble to blade tool industries in the Late Pleistocene of Northeast Asia [@Brantingham_Perreault_2010; @Gao_Zhang_Yang_Shen_Wu_2010; @pei2012shuidonggou]. Shuidonggou Localities SDG 2, 7-12 have blades and bladelets made of higher quality or exotic raw materials in addition to more than 80 polished ostrich egg-shell beads (SDG 2), bone needles and an awl (SDG 12) [@pei2012shuidonggou]. The blade compoenent is claimed to have come from Siberia and/or Mongolia at 41 ka, and at 10.5 ka  microblade technology appears but the origin is uncertain. <!-- expand on ages of all the things, comment about the cause of the new things at this site, check, https://link.springer.com/article/10.1007/s10963-019-09129-w  what is the debate? -->

[@yi2014significance; @gao2014discovery; @zhang2016bone]
